Estimating Information Transfer from Different Sources in High-dimensional Linear Regression

### 2021-07-22 revision_journal/setup.tex
intro wip

We describe the data model and the underlying assumptions.
Suppose we have two regression datasets (or tasks).
We denote their sample sizes by $n_1$ and $n_2$, respectively.
Consider either $i = 1$ or $i = 2$.
Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates.
We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for the feature covariates $X^{(i)}$ follows a linear model plus random noise $\varepsilon^{(i)}$:
$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)}.$
Both $\beta^{(1)}$ and $\beta^{(2)}$ are two arbitrary deterministic or random vectors that are independent of the feature covariates and the noise vector.
Throughout the paper, we make the following assumptions on  $X^{(i)}$ and  $\varepsilon^{(i)}$, which are standard in recent  random matrix theory literature (see e.g. \citet{tulino2004random,bai2010spectral}). %We remark that all these assumptions are natural in high-dimensional statistics.

First, we assume the row vectors of $X^{(i)}$ are i.i.d. centered random vectors with  population covariance matrix $\Sigma^{(i)} \in \real^{p \times p}$. More precisely, let
\be\label{XofZ}
    X^{(i)} = Z^{(i)} (\Sigma^{(i)})^{1/2} \in \real^{n_i\times p},
\ee
where $\Sigma^{(i)}$ is a deterministic positive definite symmetric matrices, $Z^{(i)}=(z^{(i)}_{j,k})$ is a $n_i\times p$ random matrix with real-valued independent entries having zero mean and unit variance.
%\begin{equation}\label{assm1}
%\mathbb{E} z^{(1)}_{ij}=\mathbb{E} z^{(2)}_{ij} =0, \ \quad \ \mathbb{E} \vert z^{(1)}_{ij} \vert^2=\mathbb{E} \vert z^{(2)}_{ij} \vert^2 =1,
%\end{equation}
Let $\tau>0$ be a small constant.
We assume that for some constant $\varphi>4$, the $\varphi$-th moment of each entry $z^{(i)}_{j, k}$ is bounded from above by $1/\tau$:
\be \label{conditionA2}
\ex{\vert z^{(i)}_{j, k} \vert^\varphi}  \le \frac 1{\tau}. %,\quad \mathbb{E} \vert  z^{(2)}_{ij} \vert^\varphi   \le \tau^{-1}
\ee
%$Denote by $\Sig^{(i)} = O_i\Lambda_i O_i^\top$ its eigendecomposition, where $\Lambda_1=\text{diag}(\si_1^{(i)}, \ldots, \si^{(i)}_n)$.
%\be\label{eigen}
%\Sig^{(i)}, \quad  ,  \quad i=1,2,
%\ee
Additionally, We assume that the eigenvalues of $\Sigma^{(i)}$ are bounded between $\tau$ and $1 / \tau$.
%\begin{equation}\label{assm3}
%    \tau \le  \si^{(i)}_p \le\cdots\le \si^{(i)}_2 \le \si^{(i)}_1 \le \tau^{-1} \quad i=1,2.
%\end{equation}

@@

Finally, we assume a high-dimensional setting when the sample sizes are proportional to the dimension up to a constant factor.
Denote by $\rho_i = n_i / p$, we require that
\be\label{assm2}
1+\tau \le \rho_i \le p^{\tau^{-1}}.
\ee
The first inequality ($1 + \tau \le \rho_i$) ensures that the sample covariance matrices $(X^{(1)})^\top X^{(1)}$ and $(X^{(2)})^\top X^{(2)}$ are non-singular with high probability.
%Thus, the ordinary least squares estimator is well-defined for the linear regression problem on each (single) task.
%This setting encompasses many 
The second inequality ($\rho_i \le p^{\tau^{-1}}$) is without loss of generality.
Otherwise, standard concentration results such as the central limit theorem already imply accurate estimates of the linear model under the setting defined by equation \eqref{assm2}.

@@

Having defined the data model for each task, next, we describe typical scenarios in transfer learning under our setup.
We define two settings as follows:
\begin{itemize}
    \item \textbf{Covariate shift:} independent $X^{(1)}$ and $X^{(2)}$ with the same population covariance matrices but different models.
    \item \textbf{Model shift:} independent $X^{(1)}$ and $X^{(2)}$ with the same population covariance matrices (i.e. $\Sigma^{(1)}=\Sigma^{(2)}$) but different models (i.e. $\beta^{(1)} \neq \beta^{(2)}$).
\end{itemize}
One interesting special case of capturing different models is the following random-effect model, which has been studied in several recent works concerning high-dimensional (single-task) ridge regression \cite{dobriban2018high} and distributed regression \cite{dobriban2020wonder}.
 %{How does hard parameter sharing scale with sample sizes and covariate shift $M$?} One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift. Next, we illustrate how varying them impact the prediction loss.
%\begin{example}%[Sample sizes]
%\label{ex_sample_ratio}
	%We first consider the impact of varying sample sizes.
	%Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.
	%We again consider the random effect model in Example \ref{ex_same_cov} with $t=2$.
Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, let
\begin{align}
    \beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, \label{eq_re}
\end{align}
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} \mu^2$.

@@

\subsection{Estimators}

%We describe our estimators that combine the information from both datasets.
We learn a two-layer linear neural network with parameters $A \in \real^{p\times r}$, $B_1 \in \real^{r}$, and $B_2 \in \real^r$ by minimizing the following optimization objective:
\begin{align}\label{eq_hps}
    f(A, B_1, B_2) = \bignorm{X^{(1)} A B_1 - Y^{(1)}}_2^2 + \bignorm{X^{(2)} A B_2 - Y^{(2)}}_2^2.
\end{align}
The above two-layer neural network (also called hard parameter sharing \cite{C97}) captures the intuition that both datasets share the feature space (i.e. the subspace spanned by the columns of $A$) while each having a separate prediction head.
Suppose that $(\hat{A}, \hat{B_1}, \hat{B_2})$ is a minimizer of $f(\cdot)$.
We define the hard parameter sharing estimator for task $i$ as
\[ \hat{\beta}_i^{\MTL} \define \hat{B} \hat{A}_i. \]
Our study of such estimators is motivated by the fact that hard parameter sharing is widely used in empirical studies of multi-task learning \cite{R17}.
Later on in Section \ref{sec_diff}, we will evaluate the performance of HPS estimators against several natural transfer learning estimators.

@@

@@

\paragraph{Connection to information transfer.} To be precise, we say that task one provides a \textit{positive transfer} to task two, if the excess risk $L(\hat{\beta}_2^{\MTL})$ is smaller than $L(\hat{\beta}_2^{\STL})$.
Otherwise, we say that task one provides a \textit{negative transfer} to task two.
We now illustrate that the bias-variance decomposition of Proposition \ref{lem_HPS_loss} characterizes whether or not the information transfer is positive.

First, we compare the bias. Since $n_1 \ge (1 + \tau) p$ by Assumption \ref{assm_big1}, the bias of $\hat{\beta}_2^{\STL}$ is zero.
Thus, the bias of $\hat{\beta}_2^{\MTL}$ is always higher than that of $\hat{\beta}_2^{\STL}$.
Second, we compare the variance. By a similar argument as in Proposition \ref{lem_HPS_loss}, the variance of $\hat{\beta}_2^{\STL}$ is equal to $\sigma^2 \bigtr{\Sigma^{(2)} ((X^{(2)})^{\top} X^{(2)})^{-1}}$.
Then, a straightforward algebraic calculation (e.g. using the Woodbury matrix identity) shows that the variance of $\hat{\beta}_2^{\MTL}$ is always higher than that of $\hat{\beta}_2^{\STL}$:
\begin{align*}
    \bigtr{\Sigma^{(2)} \big(\hat{\Sigma}(a)\big)^{-1}} \le \bigtr{\Sigma^{(2)} \big((X^{(2)})^{\top} X^{(2)}\big)^{-1}}.
\end{align*}
Combining both comparisons, we conclude that HPS affects the excess risk in two ways: the bias always increases while the variance always decreases.
Thus, whether or not task one provides a positive transfer to task two depends on which effect is more significant.

>>>>

%We describe the data model and the underlying assumptions.
Recall we have two regression datasets (or tasks) with sample sizes denoted as $n_1$ and $n_2$, respectively.
Suppose that $i = 1$ or $i = 2$.
Let $X^{(i)} \in \real^{n_i \times p}$ be the matrix notation corresponding to dataset $i$'s covariates.
Let $Y^{(i)} \in \real^{n_i}$ and $\varepsilon \in \real^{n_i}$ be the vector notation corresponding to its labels and additive noise.
Equation \eqref{eq_linear} thus implies that
$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)}.$
Both $\beta^{(1)}$ and $\beta^{(2)}$ are two arbitrary deterministic or random vectors that are independent of the feature covariates and the noise vector.
Throughout the paper, we make the following assumptions on  $X^{(i)}$ and  $\varepsilon^{(i)}$, which are standard in the random matrix theory literature (see e.g., \citet{tulino2004random,Bai2006}). %We remark that all these assumptions are natural in high-dimensional statistics.

First, we assume the row vectors of $X^{(i)}$ are i.i.d. centered random vectors with  population covariance matrix $\Sigma^{(i)} \in \real^{p \times p}$. Equation \eqref{eq_rvm} implies that
\be\label{XofZ}
    X^{(i)} = Z^{(i)} (\Sigma^{(i)})^{1/2} \in \real^{n_i\times p},
\ee
recalling that $\Sigma^{(i)}$ is a deterministic positive definite symmetric matrices, $Z^{(i)}=(z^{(i)}_{j,k})$ is a $n_i\times p$ random matrix with real-valued independent entries having zero mean and unit variance.
%\begin{equation}\label{assm1}
%\mathbb{E} z^{(1)}_{ij}=\mathbb{E} z^{(2)}_{ij} =0, \ \quad \ \mathbb{E} \vert z^{(1)}_{ij} \vert^2=\mathbb{E} \vert z^{(2)}_{ij} \vert^2 =1,
%\end{equation}
Let $\tau>0$ be a small constant.
We assume that for some constant $\varphi>4$, the $\varphi$-th moment of each entry $z^{(i)}_{j, k}$ is bounded from above by $1/\tau$:
\be \label{conditionA2}
\ex{\vert z^{(i)}_{j, k} \vert^\varphi}  \le \frac 1{\tau}. %,\quad \mathbb{E} \vert  z^{(2)}_{ij} \vert^\varphi   \le \tau^{-1}
\ee
%$Denote by $\Sig^{(i)} = O_i\Lambda_i O_i^\top$ its eigendecomposition, where $\Lambda_1=\text{diag}(\si_1^{(i)}, \ldots, \si^{(i)}_n)$.
%\be\label{eigen}
%\Sig^{(i)}, \quad  ,  \quad i=1,2,
%\ee
Additionally, We assume that the eigenvalues of $\Sigma^{(i)}$ are bounded between $\tau$ and $1 / \tau$.
%\begin{equation}\label{assm3}
%    \tau \le  \si^{(i)}_p \le\cdots\le \si^{(i)}_2 \le \si^{(i)}_1 \le \tau^{-1} \quad i=1,2.
%\end{equation}

@@

Finally, we assume a high-dimensional setting when the sample sizes are proportional to the dimension up to a constant factor.
Denote by $\rho_i = n_i / p$, we require that
\be\label{assm2}
    \tau \le \rho_1 \le p^{\tau^{-1}} \text{ and } 1+\tau \le \rho_2 \le p^{\tau^{-1}}.
\ee
The first inequality ($1 + \tau \le \rho_i$) ensures that the sample covariance matrices $(X^{(1)})^\top X^{(1)}$ and $(X^{(2)})^\top X^{(2)}$ are non-singular with high probability.
%Thus, the ordinary least squares estimator is well-defined for the linear regression problem on each (single) task.
%This setting encompasses many 
The second inequality ($\rho_i \le p^{\tau^{-1}}$) is without loss of generality.
Otherwise, standard concentration results such as the central limit theorem already imply accurate estimates of the linear model under the setting defined by equation \eqref{assm2}.

@@

We will consider a random-effect model as our running example, which has been studied in several recent works concerning high-dimensional (single-task) ridge regression \cite{dobriban2018high} and distributed regression \cite{dobriban2020wonder}.
Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, let
\begin{align}
    \beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, \label{eq_re}
\end{align}
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $\mu^2 / p$.

@@

\subsection{Risks}

@@

\paragraph{Information transfer.} To be precise, we say that task one provides a \textit{positive transfer} to task two, if the excess risk $L(\hat{\beta}_2^{\MTL})$ is smaller than $L(\hat{\beta}_2^{\STL})$.
Otherwise, we say that task one provides a \textit{negative transfer} to task two.

@@

We now illustrate that the bias-variance decomposition of Proposition \ref{lem_HPS_loss} characterizes whether or not the information transfer is positive.
First, we compare the bias. Since $n_1 \ge (1 + \tau) p$ by Assumption \ref{assm_big1}, the bias of $\hat{\beta}_2^{\STL}$ is zero.
Thus, the bias of $\hat{\beta}_2^{\MTL}$ is always higher than that of $\hat{\beta}_2^{\STL}$.
Second, we compare the variance. By a similar argument as in Proposition \ref{lem_HPS_loss}, the variance of $\hat{\beta}_2^{\STL}$ is equal to $\sigma^2 \bigtr{\Sigma^{(2)} ((X^{(2)})^{\top} X^{(2)})^{-1}}$.
Then, a straightforward algebraic calculation (e.g. using the Woodbury matrix identity) shows that the variance of $\hat{\beta}_2^{\MTL}$ is always higher than that of $\hat{\beta}_2^{\STL}$:
\begin{align*}
    \bigtr{\Sigma^{(2)} \big(\hat{\Sigma}(a)\big)^{-1}} \le \bigtr{\Sigma^{(2)} \big((X^{(2)})^{\top} X^{(2)}\big)^{-1}}.
\end{align*}
Combining both comparisons, we conclude that HPS affects the excess risk in two ways: the bias always increases while the variance always decreases.
Thus, whether or not task one provides a positive transfer to task two depends on which effect is more significant.
